<!--
This is for including Chapter 1.  Notice that it's also good practice to name your chunk.  This will help you debug potential issues as you knit.  The chunk above is called intro and the one below is called chapter1.  Feel free to change the name of the Rmd file as you wish, but don't forget to change it here from chap1.Rmd.
-->

<!--
The {#rmd-data} text after the chapter declaration will allow us to link throughout the document back to the beginning of Chapter 1.  These labels will automatically be generated (if not specified) by changing the spaces to hyphens and capital letters to lowercase.  Look for the reference to this label at the beginning of Chapter 2.
-->

```{r wrap-hook, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```


```{r load_pkgs, echo=FALSE, message=FALSE, warning=FALSE}
# List of packages required for this analysis
pkg <- c("dplyr", "sf", "ggplot2","rnaturalearth", "rnaturalearthdata", 
         "ggspatial", "kableExtra", "ncdf4", "stars", "magick", "RcmdrMisc",
         "knitr", "bookdown", "devtools")
# Check if packages are not installed and assign the
# names of the packages not installed to the variable new.pkg
new.pkg <- pkg[!(pkg %in% installed.packages())]
# If there are any packages in the list that aren't installed,
# install them
if (length(new.pkg))
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
# Load packages (thesisdown will load all of the packages as well)
library(thesisdown)
library(dplyr)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(knitr)
library(kableExtra)
library(ncdf4)
library(stars)
library(magick)
library(RcmdrMisc)
library(ggrepel)
library(grid)
library(gridExtra)
library(cowplot)
library(xts)
```


```{r loadstations, include=FALSE}
#Load IDEAM and ISD Stations
con1 = src_postgres(dbname = "winddata", host = "localhost", port = 5432, 
                    user = "user1", password = "user1")

#Get Ideam Stations Table
originalfields4 = c("objectid", "codigo1", "nombre", "latitud", 
                    "longitud", "categoria")
originalfields4 = paste(originalfields4, collapse= ", ", sep = "")
query4 = paste("select", originalfields4, "from ideam_all_stations", 
               "where inpqrs2 = 'YES'", sep=" ")
ideam_stations = as_tibble(tbl(con1, sql(query4)))
Encoding(ideam_stations$categoria) <- "UTF-8"
Encoding(ideam_stations$nombre) <- "UTF-8"

originalfields3 = c("id", "usaf", "station_name", "latitud", "longitud")
originalfields3 = paste(originalfields3, collapse= ", ", sep = "")
query3 = paste("select", originalfields3, 
        "from isd_all_stations where usaf_isd_dataua != ''", sep=" ")
isd_stations = as_tibble(tbl(con1, sql(query3)))
```

# Data {#rmd-data}

Input data is made up of three different sources a) IDEAM - Institute of Hydrology, Meteorology and Environmental Studies of Colombia <http://www.ideam.gov.co>, b) ISD - Integrated Surface Database <https://www.ncdc.noaa.gov/isd>, and c) ERA5 climate reanalysis <https://www.ecmwf.int/en/forecasts/datasets/reanalysis-datasets/era5>. 

```{r tabledatasources1, echo= F, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
data_sources_thesis_summary1 <-read.csv("./data/data_sources_thesis_summary1.csv",header=TRUE, stringsAsFactors = F)

kable(data_sources_thesis_summary1, align=rep('l', 3),
      #col.names = c("Name[Code]", "Latitud", "Longitud"),
      caption = "Datasets description",
      caption.short = "Datasets",
      longtable = TRUE,
      booktabs = TRUE) %>%
      row_spec(0, align = "l") %>%
      column_spec(2,width = "0.8in") %>%
      column_spec(3,width = "4in") %>%
kable_styling(font_size = 10, latex_options="scale_down")

```

```{r tabledatasources2, echo= F, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
data_sources_thesis_summary2 <-read.csv("./data/data_sources_thesis_summary2.csv",header=TRUE, stringsAsFactors = F)

kable(data_sources_thesis_summary2, align=rep('l', 3),
      #col.names = c("Name[Code]", "Latitud", "Longitud"),
      caption = "Datasets variables",
      caption.short = "Variables",
      longtable = TRUE,
      booktabs = TRUE) %>%
      row_spec(0, align = "l") %>%
      column_spec(2,width = "1.2in") %>%
      column_spec(3,width = "3.5in") %>%
kable_styling(font_size = 10, latex_options="scale_down")

```

```{r tabledatasources3, echo= F, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
data_sources_thesis_summary3 <-read.csv("./data/data_sources_thesis_summary3.csv",header=TRUE, stringsAsFactors = F)

kable(data_sources_thesis_summary3, align=rep('l', 4),
      #col.names = c("Name[Code]", "Latitud", "Longitud"),
      caption = "Variables units and time",
      caption.short = "Units and Time",
      longtable = TRUE,
      booktabs = TRUE) %>%
      row_spec(0, align = "l") %>%
      column_spec(2,width = "0.8in") %>%
      column_spec(3,width = "3in") %>%
kable_styling(font_size = 10, latex_options="scale_down")
```

Ideal data source to create extreme wind speeds maps should be field observed data from IDEAM, but there are not enough number of stations around the study area to represent all the local wind variability in a huge country with multiple variety of climates and and changing thermal floors, but there are other important motivation to include different sources trying to improve output results:

* As just mentioned, low quantity of IDEAM stations
* There are uncertainties related to the way IDEAM anemometers are registering data, then comparison with other data sources are needed to be able to do appropriate data standardization, needed as a prerequisite to the analysis.
* There is no time continuity in the registration of IDEAM data. Historical time series are different and variable in each station.

Importance of ISD database for this study is based on the fact that post-processed ISD database has wind extreme values, and it was used to create extreme wind maps for United States. ISD allows comparison with IDEAM records to take better decisions in order to do needed data standardization.

Despite that ERA5 data are not observed data, but forecast, its main advantage is data availability to assess the local climatic variance every 33 square kilometers.


## IDEAM

Historical observed wind speeds from `r length(ideam_stations$codigo1)` stations in Colombia are managed by the official environmental authority IDEAM. Table \@ref(tab:tableideamstations) shows a sample of five IDEAM stations. Figure \@ref(fig:plotideamstations) shows a map of IDEAM stations.

```{r tableideamstations, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
tableideam =  ideam_stations[c(81,120,200,57,179), c("nombre", "latitud", "longitud")]
tableideam$latitud = round(tableideam$latitud, 2)
tableideam$longitud = round(tableideam$longitud, 2)

#tableideamstations = 
kable(tableideam, 
      col.names = c("Name[Code]", "Latitud", "Longitud"),
      caption = "IDEAM Stations sample",
      caption.short = "IDEAM Stations",
      longtable = TRUE,
      booktabs = TRUE)

#kable_styling(tableideamstations, font_size = 10)

```

```{r sfideam, echo=FALSE, message=FALSE, warning=FALSE}
#Create simple features from Ideam Stations
ideam_stations = st_as_sf(ideam_stations, coords = c("longitud", "latitud"), 
                          crs = 4326)
```

```{r plotideamstations, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="IDEAM Stations. Colombia"}
#Plot IDEAM Stations
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))


colombia = world_points$name == "Colombia"
panama = world_points$name == "Panama"
peru= world_points$name == "Peru"
brazil= world_points$name == "Brazil"
venezuela= world_points$name == "Venezuela"
ecuador= world_points$name == "Ecuador"

ggplot(data = world) + 
  geom_sf(fill= "antiquewhite") + 
  geom_text(data= world_points[venezuela,],aes(x=-67, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-79.2, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-78.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-67, y=-2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 13, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = ideam_stations, size=1, aes(shape=categoria, 
   color=categoria), show.legend = "point") + 
  scale_color_discrete(name = 'Category', labels = c("Agrometeorological", 
   "Ordinary Climatic", "Main Climatic", "Mareographic", 
   "Special Meteorological", "Main Synoptic")) +
  scale_shape_discrete(name = 'Category', labels = c("Agrometeorological", 
   "Ordinary Climatic", "Main Climatic", "Mareographic", 
   "Special Meteorological", "Main Synoptic")) +  
  annotation_scale(location = "bl", width_hint = 0.5) + 
  annotation_north_arrow(location = "br", which_north = "true", 
   pad_x = unit(0.05, "in"), pad_y = unit(0.05, "in"), 
   style = north_arrow_fancy_orienteering) + 
  coord_sf(xlim = c(-82.1, -63.8), ylim = c(-7.5, 15.5), expand = FALSE) +
  xlab("Longitude") + 
  ylab("Latitude") + 
  ggtitle("IDEAM Stations") + 
  theme(panel.grid.major = element_line(color = gray(.5), 
   linetype = "dashed", size = 0.5), 
   panel.background = element_rect(fill = "aliceblue"))
```

Following, the time series, autocorrelation function, and partial autocorrelation function, for IDEAM station "21205791" will be displayed.

```{r plotoneideamstation, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="IDEAM Station - Time Series"}
#Plote time series of one IDEAM Station
originalfields = c("21205791")
newfields = paste ("X", originalfields, sep="")
originalfields = paste('"', originalfields, '"', sep = "")
newfields = paste('"', newfields, '"', sep = "")

fiedls_query = paste(originalfields, "as", newfields, sep = " ")
fiedls_query = c(paste('"', "mydatetime", '"', sep = ""), fiedls_query)
fiedls_query = paste (fiedls_query, "", sep= "", collapse=", ")

wherestring = c("21205791")
wherestring = paste('"', wherestring, '"', sep = "")
wherestring = paste(wherestring, "IS NOT NULL", sep = " ")
wherestring = paste(wherestring, collapse = " OR " , sep = " ")
query = paste("select", fiedls_query, "from ideam_vvmx_60", "where", 
              wherestring, sep=" ")

all_vvmx_aut_60 = as_tibble(tbl(con1, sql(query)))
timestamp_all_vvmx_aut_60 <- as.POSIXct(as_tibble(select(all_vvmx_aut_60, 
            mydatetime))$mydatetime,format="%Y-%m-%d %H:%M:%S", tz="UTC")

statideam_xts = na.omit(xts(x=select(all_vvmx_aut_60, "X21205791"), 
                            order.by = timestamp_all_vvmx_aut_60))
par(oma = c(2,0,0,0))
plot.xts(statideam_xts, main = "Station ID: 21205791\nWind Velocity [m/s]", 
         major.ticks="years", format.labels = "%b-%d\n%Y", 
         legend.loc = "top", col="green", cex.main=0.2)
mtext(side = 1, text = "Station Name: ELDORADO CATAM - AUT", outer = TRUE)
```

```{r plotoneideamstationacf, include=FALSE, fig.cap="IDEAM Station ACF"}
# par(oma = c(0,0,0,0))
# myacf = acf(statideam_xts[,1][!is.na(statideam_xts[,1])], plot=FALSE)
# titleacf = paste("Autocorrelation Function - ACF", paste("IDEAM", "21205791", sep = ":"), sep = "\n")
# plot(myacf, main = titleacf)
```

```{r plotoneideamstationpacf, include=FALSE, fig.cap="IDEAM Station PACF"}
# par(oma = c(0,0,0,0))
# mypacf = pacf(statideam_xts[,1][!is.na(statideam_xts[,1])], plot=FALSE)
# titlepacf = paste("Partial Autocorrelation Function - PACF", paste("IDEAM", "21205791", sep = ":"), sep = "\n")
# plot(mypacf, main = titlepacf)
```

## ISD

ISD is a database with environmental variables among then extreme wind speeds. ISD has data for the whole planet, and is based on observed data at meteorological stations in each country, which means that for Colombia is based on IDEAM data. Main advantage is data availability at neighbor countries and specialized post-processing made by NOAA's National Centers for Environmental Information - NCEI in United States, which facilitates its use. Table \@ref(tab:tableisdstations) shows a sample of five ISD stations. Figure \@ref(fig:plotisdstations) shows a map of ISD stations.

```{r tableisdstations, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
tableisd =  isd_stations[c(81,20,29,57,60), c("usaf", "station_name", "latitud", "longitud")]
tableisd$latitud = round(tableisd$latitud, 2)
tableisd$longitud = round(tableisd$longitud, 2)

#tableideamstations = 
kable(tableisd, 
      col.names = c("Code", "Name", "Latitud", "Longitud"),
      caption = "ISD Stations sample",
      caption.short = "ISD Stations",
      longtable = TRUE,
      booktabs = TRUE)

#kable_styling(tableideamstations, font_size = 10)
```

```{r isdsf, echo=FALSE, message=FALSE, warning=FALSE}
#Create simple features from IDEAM stations
isd_stations = st_as_sf(isd_stations, coords = c("longitud", "latitud"), 
 crs = 4326)
```


```{r plotisdstations, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="ISD Stations. Colombia and surroundings"}
#Plot ISD Stations
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

colombia = world_points$name == "Colombia"
panama = world_points$name == "Panama"
peru= world_points$name == "Peru"
brazil= world_points$name == "Brazil"
venezuela= world_points$name == "Venezuela"
ecuador= world_points$name == "Ecuador"

ggplot(data = world) + 
  geom_sf(fill= "antiquewhite") + 
  geom_text(data= world_points[venezuela,],aes(x=-67, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-79.2, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-78.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-67, y=-2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 13, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = isd_stations, size=1, aes(color= "ISD Stations"), shape=2, 
   show.legend = "point") + 
  scale_color_manual(values = c("ISD Stations" = "black"), name="") +
  annotation_scale(location = "bl", width_hint = 0.5) + 
  annotation_north_arrow(location = "br", which_north = "true", 
   pad_x = unit(0.05, "in"), pad_y = unit(0.05, "in"), 
   style = north_arrow_fancy_orienteering) + 
  coord_sf(xlim = c(-82.1, -63.8), ylim = c(-7.5, 15.5), expand = FALSE) +
  xlab("Longitude") + 
  ylab("Latitude") + 
  ggtitle("Integrated Surface Database - ISD") + 
  theme(panel.grid.major = element_line(color = gray(.5), 
   linetype = "dashed", size = 0.5), 
   panel.background = element_rect(fill = "aliceblue"))
```

Following, the time series, autocorrelation function, and partial autocorrelation function, for ISD station "802590" will be displayed.

```{r plotoneisdstation, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="ISD Station - Time Series"}
#Plot one ISD Station
originalfields1 = c("802590")
newfields1 = paste ("X", originalfields1, sep="")
originalfields1 = paste('"', originalfields1, '"', sep = "")
newfields1 = paste('"', newfields1, '"', sep = "")
fiedls_query1 = paste(originalfields1, "as", newfields1, sep = " ")
fiedls_query1 = c(paste('"', "mydatetime", '"', sep = ""), fiedls_query1)
fiedls_query1 = paste (fiedls_query1, "", sep= "", collapse=", ")

wherestring1 = c("802590")
wherestring1 = paste('"', wherestring1, '"', sep = "")
wherestring1 = paste(wherestring1, "IS NOT NULL", sep = " ")
wherestring1 = paste(wherestring1, collapse = " OR " , sep = " ")
query1 = paste("select", fiedls_query1, "from isd_lite_unstack", "where", 
 wherestring1, sep=" ")

isdlite = as_tibble(tbl(con1, sql(query1)))

timestamp_isdlite <- as.POSIXct(as_tibble(select(isdlite, 
 mydatetime))$mydatetime,format="%Y-%m-%d %H:%M:%S", tz="UTC")

statisd_xts = na.omit(xts(x=select(isdlite, "X802590"), 
 order.by = timestamp_isdlite))
par(oma = c(2,0,0,0))
plot.xts(statisd_xts, main = "Station ID: 802590\nWind Velocity [m/s]", 
 major.ticks="years", format.labels = "%b-%d\n%Y", legend.loc = "top", 
 col="green", cex.main=0.2)
mtext(side = 1, text = "Station Name: ALFONSO BONILLA ARAGON INTL", 
 outer = TRUE)
```

```{r plotoneisdstationacf, include=FALSE, fig.cap="ISD Station ACF"}
# par(oma = c(0,0,0,0))
# myacf = acf(statisd_xts[,1][!is.na(statisd_xts[,1])], plot=FALSE)
# titleacf = paste("Autocorrelation Function - ACF", paste("ISD", "802590", sep = ":"), sep = "\n")
# plot(myacf, main = titleacf)
```

```{r plotoneisdstationpacf, include=FALSE, fig.cap="IDEAM Station PACF"}
# par(oma = c(0,0,0,0))
# mypacf = pacf(statisd_xts[,1][!is.na(statisd_xts[,1])], plot=FALSE)
# titlepacf = paste("Partial Autocorrelation Function - PACF", paste("ISD", "802590", sep = ":"), sep = "\n")
# plot(mypacf, main = titlepacf)
```


## ERA5

```{r loadera5, include=FALSE}
ncname <- "outfile_nc4c_zip9"
filename <- paste("./data/", ncname, ".nc", sep = "")
ncin <- nc_open(filename)
lon <- ncvar_get(ncin, "longitude")
nlon = dim(lon)
lat <- ncvar_get(ncin, "latitude")
nlat = dim(lat)
ntime <-  dim(ncvar_get(ncin, "time"))
variablename <- "fg10"
fg10.units <- ncatt_get(ncin, variablename, "units")
fg10.units
lonlat.unstack <- expand.grid(lon=as.numeric(lon), lat=as.numeric(lat))
```


```{r era5sf, include=FALSE}
#Create ERA5 centers (point with lat, lon, and value, as cell index)
era5colpoints = st_as_sf(lonlat.unstack, coords=1:2, crs=st_crs(4326))
era5colpoints$value = 1:(nlon*nlat)
```

```{r era5bbox, include=FALSE}
#Define stars object to match with ERA5 bounding box.
#Cell centers of stars object, need to be same cell centers of ERA5
pointsbbox = st_bbox(era5colpoints)
cellsize = lonlat.unstack$lon[2]- lonlat.unstack$lon[1]
mybbox = st_bbox(c(pointsbbox$xmin - (cellsize/2), pointsbbox$xmax + 
 (cellsize/2), pointsbbox$ymax + (cellsize/2), pointsbbox$ymin - 
 (cellsize/2)), crs = st_crs(4326))
era5colraster.st = st_rasterize(era5colpoints, st_as_stars(mybbox, 
 nx = nlon, ny = nlat, values = era5colpoints$value))
```


ERA5 is forecast reanalysis data processed by the _European Center for Medium-Range Weather Forecasts_ - ECMWF with wind speeds time series in square cells _matrix of pixels_ of 0.25 degrees (33 km) covering the whole planet. For the study area was extracted a raster of `r (nlat)` rows by `r nlon` columns in format netCDF. Figure \@ref(fig:plotera5stations) shows a map of ERA5 stations (cells centers).


```{r era5pols, message=FALSE, warning=FALSE, include=FALSE}
#Load ERA5 polygon vectors, representing cells of ERA5
file_era5_sf_pol = "./data/era5grid_left_right_pol.shp"
era5_4326_sf_pol = st_read(dsn=file_era5_sf_pol)
pts <- do.call(rbind, st_centroid(st_geometry(era5_4326_sf_pol)))
x = pts[,1]
y = pts[,2]
era5_4326_sf_pol$x = x
era5_4326_sf_pol$y = y
era5_4326_sf_pol_filter_corners = era5_4326_sf_pol %>% 
 filter(DN %in% c(1, 49, 3333, 3381))
era5_4326_sf_pol_filter_corners_left = era5_4326_sf_pol %>% 
 filter(DN %in% c(1, 3333))
era5_4326_sf_pol_filter_corners_right = era5_4326_sf_pol %>% 
 filter(DN %in% c(49, 3381))
```


```{r plotera5stations, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="ERA5 Cells and Stations (cells centers). 49 cols by 69 rows. Cell size 0.25 decimal degrees (aprox 28 km in Colombia). Station IDs from 1 (lon=-79, lat=12.5) to 3381 (lon=-67, lat=-4.5)", fig.height=4.05}
#Plot ERA5 Stations
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

colombia = world_points$name == "Colombia"
panama = world_points$name == "Panama"
peru= world_points$name == "Peru"
brazil= world_points$name == "Brazil"
venezuela= world_points$name == "Venezuela"
ecuador= world_points$name == "Ecuador"

big = ggplot(data = world) + 
  geom_sf(fill= "antiquewhite") +
  geom_sf(data=era5colpoints, size=0.1, aes(color = "Stations"), 
   shape=".",  show.legend = "point")+
  scale_color_manual(values = c("Stations" = "black"), name="ERA5", 
   guide = guide_legend(override.aes = list(fill= c(NA), 
   linetype = c("blank"), shape = c(".")))) +
  geom_sf(data = era5_4326_sf_pol_filter_corners, color = "black", 
   aes(fill="Cells"), size=0.1, alpha=1, show.legend = "polygon") +
  scale_fill_manual(values = c("Cells" = NA), name="", 
   guide = guide_legend(override.aes = list(fill = c(NA), 
   shape = c(NA), size=0.1))) +
  geom_rect(mapping=aes(xmin=-79.252968100, xmax=-78.247031900, 
   ymin=11.832846362, ymax=12.667153638), color="red", alpha=0, 
   size=0.1) +
  geom_rect(mapping=aes(xmin=-67.752968100, xmax=-66.747031900, 
   ymin=11.832846362, ymax=12.667153638), color="red", alpha=0, 
   size=0.1) +
  geom_rect(mapping=aes(xmin=-79.258632089, xmax=-78.241367911, 
   ymin=-4.671851259, ymax=-3.828148741), color="red", alpha=0, 
   size=0.1) +
  geom_rect(mapping=aes(xmin=-67.758632089, xmax=-66.741367911, 
   ymin=-4.671851259, ymax=-3.828148741), color="red", alpha=0, 
   size=0.1) +
  geom_text(data= world_points[venezuela,],aes(x=-66.3, y=8.5, 
   label=name), color = "darkblue", fontface = "bold", size=2, 
   check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-79.7, y=9.2, 
   label=name), color = "darkblue", fontface = "bold", size=2, 
   check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-79.5, y=-1.5, 
   label=name), color = "darkblue", fontface = "bold", size=2, 
   check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75.5, y=-5.2, 
   label=name), color = "darkblue", fontface = "bold", size=2, 
   check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-66, y=-2, 
   label=name), color = "darkblue", fontface = "bold", size=2, 
   check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 14, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 2) + 
  annotate(geom = "text", x = -80, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 2) +
  geom_text_repel(data = era5_4326_sf_pol_filter_corners_left, size=2, 
   aes(x=x, y=y, label = DN), direction="y", segment.size=0.1, 
   segment.color= "grey50", color="grey50", nudge_x=-1, hjust=1, 
   box.padding=0.1) +
  geom_text_repel(data = era5_4326_sf_pol_filter_corners_right, size=2, 
   aes(x=x, y=y, label = DN), direction="y", segment.size=0.1, 
   segment.color= "grey50", color="grey50", nudge_x=+1, hjust=0, 
   box.padding=0.1) +
  coord_sf(xlim = c(-81.1, -64.8), ylim = c(-5, 13), expand = FALSE) +
  xlab("") + 
  ylab("") + 
  ggtitle("ERA5 Reanalysis - Forecast") +
  theme(plot.title = element_text(size=8)) +
  theme(axis.text.x= element_text(size=7)) + 
  theme(axis.text.y= element_text(size=7)) +
  theme(panel.grid.major = element_line(color = gray(.5), 
   linetype = "dashed", size = 0.1)) +
  theme(panel.background = element_rect(fill = "aliceblue")) +
  theme(legend.title = element_text(size=8)) +
  theme(legend.text=element_text(size=8)) +
  theme(legend.key.size = unit(0.5,"line")) +
  theme(plot.margin=unit(c(0,0,0,0),"cm")) +
  #theme(plot.background = element_rect(color = "black"))+
  theme(axis.text.x = element_text(margin =  margin(t =2, b = -10))) + 
  theme(axis.text.y = element_text(margin =  margin(r =2, l = -10)))
  
corner1lt = ggplot(data = world) + 
  geom_sf(fill= "antiquewhite",  size=0.1) + 
  geom_sf(data = era5_4326_sf_pol, colour="black", fill=NA, size=0.1) +
  geom_text(data= world_points[venezuela,],aes(x=-66.5, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-80.5, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-79.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-68, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[colombia,],aes(x=-71, y=4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 14, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80.5, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = st_cast(world, "MULTILINESTRING"),  size=0.1)+
  geom_sf_text(data = era5_4326_sf_pol, aes(label = DN), size=2) +
  coord_sf(xlim = c(-79.252968100, -78.247031900), ylim = c(11.832846362, 
   12.667153638), expand = FALSE) +
  xlab("") + 
  ylab("") + 
  ggtitle("") +
  theme(panel.grid = element_blank()) + 
  theme(panel.background = element_rect(fill = "aliceblue")) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.margin=grid::unit(c(0,0.2,0,0),"cm")) + 
  theme(panel.border = element_rect(colour = "red"))+
  theme(axis.ticks.length=unit(0, "null")) +
  theme(axis.title.x=element_blank()) +
  theme(axis.title.y=element_blank()) +
  theme(plot.title = element_blank())
  
corner2rt = ggplot(data = world) + 
  geom_sf(fill= "antiquewhite",  size=0.1) + 
  geom_sf(data = era5_4326_sf_pol, colour="black", fill=NA, size=0.1) + 
  geom_text(data= world_points[venezuela,],aes(x=-66.5, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-80.5, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-79.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-68, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[colombia,],aes(x=-71, y=4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 14, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80.5, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = st_cast(world, "MULTILINESTRING"),  size=0.1)+
  geom_sf_text(data = era5_4326_sf_pol, aes(label = DN), size=2) +
  coord_sf(xlim = c(-67.752968100, -66.747031900), ylim = c(11.832846362, 
   12.667153638), expand = FALSE) +
  xlab("") + 
  ylab("") + 
  ggtitle("") +
  theme(panel.grid = element_blank()) +
  theme(panel.background = element_rect(fill = "aliceblue")) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.margin=grid::unit(c(0,0,0,0.2),"cm")) + 
  theme(panel.border = element_rect(colour = "red"))+
  theme(axis.ticks.length=unit(0, "null")) +
  theme(axis.title.x=element_blank()) +
  theme(axis.title.y=element_blank()) +
  theme(plot.title = element_blank())

corner3lb = ggplot(data = world) + 
  geom_sf(fill= "antiquewhite",  size=0.1) + 
  geom_sf(data = era5_4326_sf_pol, colour="black", fill=NA, size=0.1) + 
  geom_text(data= world_points[venezuela,],aes(x=-66.5, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-80.5, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-79.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-68, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[colombia,],aes(x=-71, y=4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 14, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80.5, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = st_cast(world, "MULTILINESTRING"),  size=0.1)+
  geom_sf_text(data = era5_4326_sf_pol, aes(label = DN), size=2) +
  coord_sf(xlim = c(-79.258632089, -78.241367911), ylim = c(-4.671851259, 
   -3.828148741), expand = FALSE) +
  xlab("") + 
  ylab("") + 
  ggtitle("") +
  theme(panel.background = element_rect(fill = "aliceblue")) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.margin=grid::unit(c(0,0.2,0,0),"cm")) + 
  theme(panel.border = element_rect(colour = "red"))+
  theme(axis.ticks.length=unit(0, "null")) +
  theme(axis.title.x=element_blank()) +
  theme(axis.title.y=element_blank()) +
  theme(plot.title = element_blank())

corner4rb = ggplot(data = world) + 
  geom_sf(fill= "antiquewhite",  size=0.1) + 
  geom_sf(data = era5_4326_sf_pol, colour="black", fill=NA, size=0.1) + 
  geom_text(data= world_points[venezuela,],aes(x=-66.5, y=8.5, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[panama,],aes(x=-80.5, y=9.2, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) + 
  geom_text(data= world_points[ecuador,],aes(x=-79.2, y=-1, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[peru,],aes(x=-75, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[brazil,],aes(x=-68, y=-6, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  geom_text(data= world_points[colombia,],aes(x=-71, y=4, label=name), 
   color = "darkblue", fontface = "bold", size=3, check_overlap = FALSE) +
  annotate(geom = "text", x = -77.5, y = 14, label = "Caribbean\nSea", 
   fontface = "italic", color = "grey22", size = 4) + 
  annotate(geom = "text", x = -80.5, y = 5, label = "Pacific\nSea", 
   fontface = "italic", color = "grey22", size = 4) +
  geom_sf(data = st_cast(world, "MULTILINESTRING"),  size=0.1)+
  geom_sf_text(data = era5_4326_sf_pol, aes(label = DN), size=2) +
  coord_sf(xlim = c(-67.758632089, -66.741367911), ylim = c(-4.671851259, 
   -3.828148741), expand = FALSE) +
  xlab("") + 
  ylab("") + 
  ggtitle("") +
  theme(panel.background = element_rect(fill = "aliceblue")) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.margin=grid::unit(c(0,0,0,0.2),"cm")) + 
  theme(panel.border = element_rect(colour = "red"))+
  theme(axis.ticks.length=unit(0, "null")) +
  theme(axis.ticks.margin=unit(0, "null")) +
  theme(axis.title.x=element_blank()) +
  theme(axis.title.y=element_blank()) +
  theme(plot.title = element_blank())

grid.arrange(big, arrangeGrob(corner1lt, corner2rt, corner3lb, corner4rb), ncol=2, widths=c(2.2,1))
```

```{r, include=FALSE}
t.units <- ncatt_get(ncin, "time", "units")
t.units
time.array <-  ncvar_get(ncin, "time")
nt = dim(time.array)
nt
head(time.array, 1)
tail(time.array, 1)
tustr <- strsplit(t.units$value, " ")
#tustr
#Our data is in Gregorian Calendar: "hours"      "since"      "1900-01-01" "00:00:00.0"
anio_mes_dia = unlist(tustr)[3]
anio_mes_dia
#Note that the dates are converted first to seconds!
library(lubridate)
timestamp = as_datetime(c(time.array*60*60),origin=anio_mes_dia)
timestamp_string <- as.character(timestamp)
#str(timestamp_string)
  statera5_xts = na.omit(xts(x= ncvar_get(ncin, 'fg10', start=c(1,1,1), count=c(1,1,ntime)), order.by = timestamp))
  colnames(statera5_xts) = paste0("ERA5_", "1")
  
```

## Data Download and Data Organization

All data sources had different mechanisms for downloading. For IDEAM, the official procedure is through a written request using the e-mail _atencionalciudadano@ideam.gov.co_, then they will provide a link to get the information. For ISD, all files are available in the ftp site ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-lite/, organized in folders by years, then a gzip file is available, with the station name in the format _ID-99999-YYYY.gz_, where ID is the USAF-ISD station identifier. For ISD, there are many files by station, one file for each year with data available. For ERA5 it is possible to make a request using a Python script, but since there is a size limit for downloading, it is neccesary to split the request, then, use console commands to create an unified netCDF file. Files with all IDEAM and ISD stations are available in the Annex \@ref(rcode) - [Resutls - Digital Files](#results). For the Python code, and the commands to join netCDF files of ERA5 data source, see the Annex \@ref(era5download) - [ERA5 Data Download and Integration](#era5download).

\clearpage
