`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->


# Resutls - Digital Files


```{r resultsstructure, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, results="asis"}
library(knitr)
library(kableExtra)
results_structure <-read.csv("./data/results_annex/results_structure.csv",header=TRUE, stringsAsFactors = F) 
#Ftp = results_structure[,3]

mylink_markdown <- paste0("[", results_structure[,1], "](", results_structure[,3], ")")
#mylink_latex <-  paste0("\\href{", results_structure[,3], "}{\verb", knitr::asis_output("\U007C"), results_structure[,1], knitr::asis_output("\U007C"), "}")
mylink_latex <-  paste0("\\href{", results_structure[,3], "}{", results_structure[,1], "}")
#mylink_latex <-  paste0("\\href{", results_structure[,3], "}{a}"
#mytable_markdown <- cbind(folder_Tree = mylink_markdown, description = results_structure[,2])

mytable_latex <- cbind(folderTree = mylink_latex, description = results_structure[,2])


kable(mytable_latex, linesep = "", align=rep('l', 2), format = "latex", escape = FALSE, #format = "markdown",
       col.names = c("Folder Tree - Ftp Links","Description"),
       caption = "Digital files in FTP site \\href{ftp://ftp.geocorp.co/windthesis/}{ftp://ftp.geocorp.co/windthesis/}. User anonymous@geocorp.co (no password).",
       caption.short = "digital files",
       longtable = TRUE,
       booktabs = TRUE) %>%
       row_spec(0, align = "l") %>%
       column_spec(1,width = "2.3in") %>%
       column_spec(2,width = "3.3in") %>%
#       column_spec(3,width = "0.5in") %>%
       kable_styling(font_size = 8, latex_options=c("scale_down")) #%>%
#       #add_header_above(c(" ", "Raw Data" = 4, "Declustered Data" = 4))

# library(dplyr)
# library(knitr)
# library(kableExtra)
# dt_url <- c("https://en.wikipedia.org/wiki/Cadillac_Fleetwood",
#             "https://www.lincoln.com/luxury-cars/continental/",
#             "http://shop.honda.com/civics.aspx",
#             "https://bringatrailer.com/2011/12/28/striking-1973-maserati-bora-4-9/")



#results_structure[,1:2] %>%
# mutate (FolderTree = cell_spec(FolderTree, "html", link = Ftp)) %>%
# kable("html", escape = FALSE, format = "markdown") %>%
# kable_styling(bootstrap_options = c("hover", "condensed"))
   
   
   
# mtcars[c(15,16,19,31),1:3] %>% 
#   mutate(model = cell_spec(row.names(.), "html", link = dt_url)) %>%
#   kable("html", escape = FALSE) %>%
#   kable_styling(bootstrap_options = c("hover", "condensed"))

```


```{r nice-tab, tidy=FALSE,echo=FALSE,message=FALSE} 
# library(dplyr)
# library(kableExtra)
# library(rmarkdown)
# library(knitr)
# df<-data.frame(name = rep("linkname",10))
# url<- c("https://en.wikipedia.org/wiki/Cadillac_Fleetwood",
#        "https://www.lincoln.com/luxury-cars/continental/",
#        "http://shop.honda.com/civics.aspx",
#        "https://bringatrailer.com/2011/12/28/striking-1973-maserati-bora-4-9/",
#        "https://en.wikipedia.org/wiki/Cadillac_Fleetwood",
#        "https://www.lincoln.com/luxury-cars/continental/",
#        "http://shop.honda.com/civics.aspx",
#        "https://bringatrailer.com/2011/12/28/striking-1973-maserati-bora-4-9/",
#        "https://en.wikipedia.org/wiki/Cadillac_Fleetwood",
#        "https://www.lincoln.com/luxury-cars/continental/")
# 
# df$name <- paste0("[", df$name, "](", url, ")")
# 
# knitr::kable(df, format = "markdown")
# 
# x <- cars[1:3,]
# urls <- c("stackoverflow.com", "stats.stackexchange.com", "homebrew.stackexchange.com")
# linkID <- c("link1", "link2", "link3")
# x$website <- paste0("[", linkID, "](http://", urls, ")")
# sections<-c("rmd-results","rmd-results","rmd-results")
# section_links<-paste0("\\\\@ref(",sections,")")
# fig <-c("ingeniarmaps","ingeniarmaps","ingeniarmaps")
# fig_links<-paste0("\\\\@ref(fig:",sections,")")
# x$my_fig = paste0("[", linkID, "](#",fig,")")
# x$my_sections<-paste0("[", linkID, "](#",sections,")")
# knitr::kable(x, caption = 'Example table', booktabs = TRUE, format = "markdown")


```

```{r rawdatastationidfitted, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, results="asis"}
library(knitr)
library(kableExtra)
rawdatastationidfitted <-read.csv("./data/results_annex/raw_data_station_id_fitted.csv",header=TRUE, stringsAsFactors = F) 


kable(rawdatastationidfitted, linesep = "", align=rep('l', 3), #format = "markdown",
       col.names = c("Excel Sheet Name","Description","Important"),
       caption = "Content of the output Excel Book 'raw\\_data\\_station\\_*\\_fitted.xlsx', where '*' is replaced by the Station ID. One file by station.",
       caption.short = "rawdatastationidfitted",
       longtable = TRUE,
       booktabs = TRUE) %>%
       row_spec(0, align = "l") %>%
       column_spec(1,width = "1.5in") %>%
       column_spec(2,width = "3in") %>%
       column_spec(3,width = "1.5in") %>%
       kable_styling(font_size = 8, latex_options=c("scale_down")) #%>%
```

```{r rawdatastationidstatistics, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, results="asis"}
library(knitr)
library(kableExtra)
rawdatastationidstatistics <-read.csv("./data/results_annex/raw_data_station_id_statistics.csv",header=TRUE, stringsAsFactors = F) 


kable(rawdatastationidstatistics, linesep = "", align=rep('l', 2), #format = "markdown",
       col.names = c("Excel Sheet Name","Description"),
       caption = "Content of the output Excel Book 'raw\\_data\\_station\\_*\\_statistics.xlsx', where '*' is replaced by the Station ID. One file by station.",
       caption.short = "rawdatastationidstatistics",
       longtable = TRUE,
       booktabs = TRUE) %>%
       row_spec(0, align = "l") %>%
       column_spec(1,width = "1.2in") %>%
       column_spec(2,width = "4.6in") %>%
       #column_spec(3,width = "1.5in") %>%
       kable_styling(font_size = 8, latex_options=c("scale_down")) #%>%
```

```{r fittedmodelidpdf, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, results="asis"}
library(knitr)
library(kableExtra)
fittedmodelidpdf <-read.csv("./data/results_annex/fittedmodel_id_pdf.csv",header=TRUE, stringsAsFactors = F) 


kable(fittedmodelidpdf, linesep = "", align=rep('l', 2), #format = "markdown",
       col.names = c("Graphic","Description"),
       caption = "Content of the output graphics PDF file 'FittedModel\\_*.pdf', where '*' is replaced by the Station ID. One file by station.",
       caption.short = "fittedmodelidpdf",
       longtable = TRUE,
       booktabs = TRUE) %>%
       row_spec(0, align = "l") %>%
       column_spec(1,width = "0.4in") %>%
       column_spec(2,width = "5.7in") %>%
       #column_spec(3,width = "1.5in") %>%
       kable_styling(font_size = 8, latex_options=c("scale_down")) #%>%
```

return_levels.csv
fitted_model_result.xlsx 

```{r returnlevels, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, results="asis"}
library(knitr)
library(kableExtra)
returnlevels <-read.csv("./data/results_annex/return_levels.csv",header=TRUE, stringsAsFactors = F) 


kable(returnlevels, linesep = "", align=rep('l', 4), #format = "markdown",
       col.names = c("Column ID", "Columns Name", "Important", "Description"),
       caption = "Content of the output Excel Book 'fitted\\_model\\_result.xlsx' (sheet pp\\_pintar). One file by dataset (ISD, ERA5).",
       caption.short = "returnlevels",
       longtable = TRUE,
       booktabs = TRUE) %>%
       row_spec(0, align = "l") %>%
       column_spec(1,width = "0.6in") %>%
       column_spec(2,width = "1.6in") %>%
       column_spec(3,width = "1in") %>%
       column_spec(4,width = "2.5in") %>%
       kable_styling(font_size = 8, latex_options=c("scale_down")) #%>%
```


# R Code


This appendix includes all of the R chunks of code that were hidden throughout the document (using the `include = FALSE` or `echo=FALSE, message=FALSE, warning=FALSE` chunks tag) to help with readability and/or setup.

**In Chapter \@ref(rmd-data):**


```{r ref.label='plotideamstations', results='hide', echo = TRUE}
```

**In Chapter \@ref(rmd-thefra):**


```{r ref.label='plotgumbelpdffunction', results='hide', echo = TRUE}
```

# The Next Appendix


